% Credits for Concrete Mathematics
% (c) Addison-Wesley, all rights reserved.
\input gkpmac

\refin bib

\pageno=632
\parindent=30pt
\hyphenation{Reich}

\beginchapter C Credits for Exercises

THE EXERCISES in this book have been drawn from many sources.
The authors have tried to trace the origins of all the problems that have
been published before,
except in cases where the exercise is so elementary that its inventor
would probably not think anything was being invented.

Many of the exercises come from examinations in Stanford's Concrete
Mathematics classes. The teaching assistants and instructors often
\g \vskip.7in The TA sessions were invaluable, I~mean really great.\vskip.3in
Keep the same instructor and the same TAs next year.\vskip.3in
Class notes \undertext{very} good and useful.\vskip.3in
I never ``got'' Stirling numbers.\g
devised new problems for those exams, so it is appropriate to
list their names here:
\begindisplay \def\preamble{&##\hfil\qquad} \openup-2pt
\sl\kern-.5pt Year&\sl\kern-.5pt Instructor&\sl\kern-.5pt Teaching Assistant(s)\cr
\noalign{\vskip2pt}
1970&Don "Knuth"&Vaughan "Pratt"\cr
1971&Don Knuth&Leo Guibas\cr
1973&Don Knuth&Henson "Graves", Louis "Jouaillec"\cr
1974&Don Knuth&Scot "Drysdale", Tom "Porter"\cr
1975&Don Knuth&Mark "Brown", Luis "Trabb Pardo"\cr
1976&Andy Yao&Mark Brown, Lyle Ramshaw\cr
1977&Andy Yao"!Yao, Andrew"&Yossi "Shiloach"\cr
1978&Frances Yao"!Yao, Frances"&Yossi Shiloach\cr
1979&Ron "Graham"&Frank "Liang", Chris "Tong", Mark "Haiman"\cr
1980&Andy Yao&Andrei "Broder", Jim "McGrath"\cr
1981&Ron Graham&Oren "Patashnik"\cr
1982&Ernst Mayr&Joan "Feigenbaum", Dave "Helmbold"\cr
1983&Ernst "Mayr"&Anna "Karlin"\cr
1984&Don Knuth&Oren Patashnik, Alex "Sch\"affer"\cr
1985&Andrei Broder&Pang "Chen", Stefan "Sharkansky"\cr
1986&Don Knuth&Arif "Merchant", Stefan Sharkansky\cr
\enddisplay
In addition, David "Klarner" (1971), Bob "Sedgewick" (1974), Leo "Guibas" (1975),
and Lyle "Ramshaw" (1979) each
contributed to the class by giving six or more guest lectures. Detailed
lecture notes
taken each year by the teaching assistants and edited by the instructors
have served as the basis of this book.

\vfill\eject
\advance\baselineskip-.35pt
\def\source#1:{\par
 \noindent\hangindent=\parindent\hbox to\parindent{\bf #1\hfil}\ignorespaces}

\rightskip0pt plus4em \spaceskip.3333em \xspaceskip.5em
\hsize=2.75in
\newif\ifright \newbox\leftcol
\output{\ifright\twocolumnout\global\rightfalse
  \else\leftcolumnsave\global\righttrue\fi}
\def\leftcolumnsave{\global\setbox\leftcol=\box255}
\def\twocolumnout{\shipout\vbox to\totheight{ % here we define one page of output
    \offinterlineskip % butt the boxes together
    \vbox to 2pc{ % this part goes on top of the regular pages
      \ifodd\pageno \rightheadline\else\leftheadline\fi
      \vfill} % this completes the \vbox to 2pc
    \hbox to35pc{\box\leftcol\hss\box255}
    \vfill
    \ifpreprint\ifinxmode\makeinxfooter\fi\fi
    }
  \advancepageno}
\input sources
\bigskip
\noindent *\thinspace Unpublished personal communication.
%\bye
\eject\bye
